---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
#### Show any code that is needed to
1. Load the data (i.e. read.csv())
```{r}
data <- read.csv("./activity.csv")
```

2. Process/transform the data (if necessary) into a format suitable for your analysis.  
**The format of the data is suitable already for my analysis.**

## What is mean total number of steps taken per day?
#### For this part of the assignment, you can ignore the missing values in the dataset.  
1. Calculate the total number of steps taken per day
```{r, results="axis"}
ttlstepsperday <- with(data, tapply(steps, date, sum)); ttlstepsperday
```

2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day.
```{r}
hist(ttlstepsperday, main = "Histogram of Total Number of Steps Taken Each Day", xlab = "Total Number of Steps Taken Each Day", ylab = "Number of Days")
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
mean(ttlstepsperday, na.rm = TRUE)
```
**The mean total number of steps taken per day is 10766.19.**
```{r}
median(ttlstepsperday, na.rm = TRUE)
```
**The median total number of steps taken per day is 10765.**


## What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).
```{r}
avgDAP <- with(data, tapply(steps, interval, mean, na.rm=TRUE))
plot(avgDAP, main = "Average Daily Activity Pattern", type = "l", xlab = "Intervals", ylab = "Number of Steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
which.max(avgDAP)
```
**The $104^{th}$ interval, a.k.a. the 835 to 845 minute interval, contains the maximum number of steps on average across all the days.**


## Imputing missing values
#### Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.  
1. Calculate and report the total number of missing values in the dataset. (i.e. the total number of rows with NAs)
```{r}
sum(is.na(data$steps))
```
**There are 2,304 rows with NAs.**

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.  
**In the code chunk below, the NAs are replaced by the means for the 5-minute intervals.**
```{r}
vavgDAP <- rep(as.vector(avgDAP), length(unique(data$date))) #then replace the corresponding NA values
head(vavgDAP)
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
Newdata <- data #Create a new dataset
ordinal.na <- which(is.na(data$steps)) #sequential number of NAs
Newdata$steps[ordinal.na] <- vavgDAP[ordinal.na] #replace NAs
head(Newdata)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
ttlstepsperdayNew <- with(Newdata, tapply(steps, date, sum))
hist(ttlstepsperdayNew, main = "Histogram of Total Number of Steps Taken Each Day (Updated)", xlab = "Total Number of Steps Taken Each Day", ylab = "Number of Days")
```

```{r}
mean(ttlstepsperdayNew)
```
**The mean total number of steps taken per day is 10766.19.**  

```{r}
median(ttlstepsperdayNew)
```
**The median total number of steps taken per day is 10766.19.**   

**The new median is slightly greater than the old one, while the mean does not change. In this case, imputting missing data causes the original distribution slightly skew leftwards.**


## Are there differences in activity patterns between weekdays and weekends?
#### For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.
1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day. 
```{r, warning=FALSE}
Newdata$day <- weekdays(as.POSIXct(Newdata$date, "%Y-%b-%d"))
Newdata$day <- factor(ifelse(Newdata$day %in% c("Saturday", "Sunday"),
                "weekend", "weekday"))
head(Newdata)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.  
```{r}
temp <- aggregate(steps ~ interval + day, data = Newdata, mean) #temp data by intervals and day
library(lattice)
xyplot(steps ~ interval | day, data = temp, type = "l",layout = c(1,2), xlab = "Interval", ylab = "Number of Steps")
```   

**The activity pattern at weekends and on weekdays are different. On weekdays people tend to walk more on early morning. At weekends people tend to walk more in the afternoon and at night.**
